---
title: "Assignment 4 - Heart rate, respiration and interpersonal coordination"
author: "Riccardo Fusaroli"
date: "August 20, 2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment 4 - Heart rate, respiration and interpersonal coordination

Physiological data (here heart rate [variability], and respiration) are increasingly popular. Historically treated as pernicious noise to be regressed out of neuro-imaging data, there is now increasing research on how these signals tell us something important about cognition and beyond being just a signal of cognitive processes also impact them in interesting ways. Advanced sport science, and the quantified self movement (closely followed by marketing and communication) have hailed continuous physiological tracking as a powerful way to access and modify attitudes, habits, and performance. Further, as team coordination (in the military, in decision processes and organizational contexts) is more and more in focus, research has attempted to measure how interpersonal coordination between physiological systems might tell us something important about e.g. emotional and cognitive coordination. See references in the reading list for more on this.

In this assignment, you will learn to:
- collect physiological data
- pre-process physiological data (and grow further your mad R skills)
- model the continuous interdependence between two signals (using a multilevel model as proxy for a dynamical system approach)
- conservatively assess the presence of coordination between to signals in a controlled context

This assignment has two parts. The first part familiarizes you with heart rate, and respiration data and their preprocessing. The second part explores how to analyze interpersonal coordination of these signals.

These are the questions you need to be able to answer at the end of the assignment (aka that you need to submit as part of the portfolio)

1) How do you preprocess heart rate and respiration data? Describe the process. If any data needs to be excluded, list the excluded data and motivate the exclusion.

2) Do you observe interpersonal coordination in heart rate and respiration? Describe your control baseline, the method used to quantify coordination, and the statistical models used to infer whether coordination was higher than in the baseline. Report the results of the models.

3) Do you observe differences in coordination between conditions? Report the models and results.

4) Is respiration coordination a likely driver of heart rate coordination? Describe how you would test for it. Bonus points if you actually run the tests and report methods and results.

N.B. to give you a bit more data I included data from previous years (Study1, Study2 and Study 3). Note that synchronouns and turn-taking are the same across both studies, but the third condition is different: in the first year it was self-paced joint reading; in the second year it was the tv-series conversation.

## Let's get started

### Exploring physiological signals

- Choose one pair (one pair, three conditions)
- Load the logs
- Produce a plot of the participants' respiration signal and a different one of the participants' HR signal.
  N.B: remember the slides: artifacts, downsampling, scaling.
  N.B. The gridExtra::grid.arrange() function allows you to display the plots side by side. E.g. grid.arrange(plot1, plot2, plot3, ncol=3). There are also smarter packages, like cowplot and ggpubr.
- Can you eye-ball which condition if any displays more physiological coordination?

### First we read one data file and identify the procedure
- Load the file
- correctly identify all columns
- plot the data
- deal with the artifacts
- downsample the dat
- Add a column for study, group, trial and condition

```{r}
# Loading the relevant libraries
install.packages("jtools")
install.packages("ggstance")
install.packages("pacman")
library(tidyverse, tidymodels)
library(ggstance)
library(jtools)
library(goeveg)
library(DescTools)
pacman::p_load(pacman, tidyverse, tidymodels, groupdata2, lmerTest, lme4, kernlab, caret, dplyr)
pacman::p_load(goeveg, DescTools, knitr, devtools, tidyr)

# Loading a data file
data <- read.csv('~/Desktop/Cognitive Science/Cognitive Science 3rd semester/Experimental Methods 3/Afleveringer/Assignment_4/data/Study1_G1_T1_Synchronous.csv')

# Plot
#Respiration 
p1 <- ggplot(data=data, aes(x=time)) +
  geom_line(aes(y= Resp1, color = "Resp1"))+
  geom_line(aes(y= Resp2, color = "Resp2"))+
  labs( x= "Time", y= "Resp")+
  theme_classic()
p1

#Heart Rate 
p2 <- ggplot(data=data, aes(x=time)) +
  geom_line(aes(y= HR1, color = "HR1"))+
  geom_line(aes(y= HR2, color = "HR2"))+
  labs( x= "Time", y= "Heart Rate")+
  theme_classic()
p2

## Remove outliers funnction
removeOuts <- function(ts,threshold){
  ts[ts > (mean(ts,na.rm=T) +
             (threshold*sd(ts,na.rm=T))) | 
       ts < (mean(ts,na.rm=T) -
             (threshold*sd(ts,na.rm=T)))] = mean(ts,na.rm=T)
  return(ts)
}
threshold=2.5 # Default value at 2.5 sds from the mean

#adding columns with the outliers removed for Respiration 
data$RespR1_noOuts <- removeOuts(data$Resp1, threshold)
data$RespR2_noOuts <- removeOuts(data$Resp2, threshold)

#adding columns for the outliers removed for Heart Rate 
data$HR1R_noOuts <- removeOuts(data$HR1, threshold)
data$HR2R_noOuts <- removeOuts(data$HR2, threshold)

# Scaling function
z_scale <- function(column){
  column_d <- (column - mean(column)) / sd(column)
}

## Scale data
data$RespR1_scaled <- z_scale(data$RespR1_noOuts)
data$RespR2_scaled <- z_scale(data$RespR2_noOuts)

data$HR1R_scaled <- z_scale(data$HR1R_noOuts)
data$HR2R_scaled <- z_scale(data$HR2R_noOuts)


# Plot raw data againt those with the outliers removed
# Respiration
p3 <- ggplot(data=data, aes(x=time)) +
  geom_line(aes(y= Resp1, color = "Resp1"))+
  geom_line(aes(y= Resp2, color = "Resp2"))+
  geom_line(aes(y= RespR1_noOuts, color = "RespR1 - no outlier"))+
  geom_line(aes(y= RespR2_noOuts, color = "RespR2"))+
  labs( x= "Time", y= "Resp")
p3


# Heart rate
p4 <- ggplot(data=data, aes(x=time)) +
  geom_line(aes(y= HR1, color = "HR1"))+
  geom_line(aes(y= HR2, color = "HR2"))+
  geom_line(aes(y= HR1R_noOuts, color = "HR1R"))+
  geom_line(aes(y= HR2R_noOuts, color = "HR2R"))+
  labs( x= "Time", y= "Resp")
  p4


# Ploting normal outlier removed data and scaled data
# Respiration
p5 <- ggplot(data=data, aes(x=time)) +
  geom_line(aes(y= RespR1_scaled, color = "RespR1S"))+
  geom_line(aes(y= RespR2_scaled, color = "RespR2S"))+
  geom_line(aes(y= RespR1_noOuts, color = "RespR1"))+
  geom_line(aes(y= RespR2_noOuts, color = "RespR2"))+
  labs( x= "Time", y= "Resp")
p5

# Heart rate
p6 <- ggplot(data=data, aes(x=time)) +
  geom_line(aes(y= HR1R_scaled, color = "HR1"))+
  geom_line(aes(y= HR2R_scaled, color = "HR2"))+
  geom_line(aes(y= HR1R_noOuts, color = "HR1R"))+
  geom_line(aes(y= HR2R_noOuts, color = "HR2R"))
  labs( x= "Time", y= "Resp")
  p6



## Downsampling the data (grouping 1000 datapoints turning time intervals into 1s)
d1 = data %>%
 group(n = 1000, method = 'greedy') %>%
 dplyr::summarise(
   time = mean(time,na.rm=T),
   HR1 = mean(HR1R_scaled,na.rm=T),
   HR2 = mean(HR2R_scaled,na.rm=T),
   Resp1 = mean(RespR1_scaled,na.rm=T),
   Resp2 = mean(RespR2_scaled,na.rm=T))

## Plot the downsampled data
p7 <- ggplot(data = d1) +
  geom_path(aes(time, Resp1, color = "P1")) +
  geom_path(aes(time, Resp2, color = "P2")) +
  labs(x = "time", y = "Resp") +
  theme(legend.position="bottom")
p7
```


## Now we are ready to go to load and pre-process all files

Go through all the files (with a function passed onto map_df), check which files should be excluded, if any, and save the pre-processed time-series

A couple of tips:
- looping is oh so slow. Making a function and using Map/Map_df is your salvation.
- each study restarts the group numbering, so you should make sure to change that (e.g. 100 * Study + Group)
- you need to make sure all the data are meaningful or something has to be removed. Plotting is your friend. E.g. "Study1_G1_T1_Synchronous" has one bad respiration signal. We could replace it with NAs


```{r}
# Defining function to load and clean all data files
data_preprocess <- function(filename, threshold = 2.5){
  
  # load the data
  data <- read_csv(filename)
  
  # remove outliers
  data$Resp1 <- removeOuts(data$Resp1, threshold)
  data$Resp2 <- removeOuts(data$Resp2, threshold)
  data$HR1 <- removeOuts(data$HR1, threshold)
  data$HR2 <- removeOuts(data$HR2, threshold)
  
  # scale variables
  data$HR1Scaled <- scale(data$HR1)
  data$HR2Scaled <- scale(data$HR2)
  data$Resp1Scaled <- scale(data$Resp1)
  data$Resp2Scaled <- scale(data$Resp2)
  
  # downsample the data
  data <- data %>% 
    group(n = 1000, method = 'greedy') %>% 
    dplyr::summarise(
      time = mean(time, na.rm=T),
      HR1 = mean(HR1Scaled, na.rm=T),
      HR2 = mean(HR2Scaled, na.rm=T),
      Resp1 = mean(Resp1Scaled, na.rm=T),
      Resp2 = mean(Resp2Scaled, na.rm=T))
  
  # create list with new variable indeces
  fileIndex <- unlist(str_extract_all(filename, "\\d"))
  # create list with condition indeces
  conditionIndex <- unlist(str_split(filename, "T*_"))
  conditionIndex <- str_remove(conditionIndex[4], ".csv")
  
  # extracting study, group, trial, condition and ID from file title
  data <- data %>%
    add_column(study = fileIndex[2]) %>% 
    add_column(group = fileIndex[3]) %>% 
    add_column(trial = fileIndex[4]) %>% 
    add_column(condition = conditionIndex) %>% 
    add_column(ID = paste(fileIndex[2], fileIndex[3], fileIndex[4],
                          conditionIndex))
  
  return(data)
}


# Run function on all files
data <- list.files(path = "/Users/bertram/desktop/folders/R/AssignmentFour/data", pattern = ".csv", full.names=T) %>% 
    purrr::map_df(data_preprocess)

# remove spaces from ID-column
data$ID <- str_replace_all(data$ID, " ", "")

# make ID-column into type factor
data$ID <- as.numeric(as.factor(data$ID))

# Save data frame
write_csv(data_list, "full_data.csv")
```

```{r}
# Loading all data
full_data <- read.csv('~/Desktop/Cognitive Science/Cognitive Science 3rd semester/Experimental Methods 3/Afleveringer/Assignment_4/dataAssignment1000.csv')

# Eye-balling for bad data 
# plots plots plots
# Study 4p plotting loop (plots two conditions at a time)
for(i in seq(from=58, to=97, by=2)) {
  pl <- full_data %>%
    filter(study==4, ID==c(i, (i+1))) %>% 
    group_by(group,trial) %>% 
    mutate(time = seq(n())) %>% 
    ggplot()+
    geom_line(aes(time,HR1, colour="blue", group=ID))+
    geom_line(aes(time,HR2, colour="red",group=ID))+
    theme_classic()+
    facet_wrap(group~trial)
  plot(pl)
}

# Filtering out only study 4 data and setting time to sequence of seconds starting from 1
data4 <- full_data %>% 
  filter(study==4) %>% 
  group_by(ID) %>%
  mutate(time = seq(n())) 



# Remove bad data (detects straight lines) - pitfall: removes 5 datapoints per person per trial
for (i in seq(NROW(data4))){
  if (i+5 > length(seq(NROW(data4)))) {
    data4$HR1[i]=NA
  } else if (mean(data4$ID[i:(i+5)]) != data4$ID[i]) {
    data4$HR1[i] = NA
  } else if ((sd(data4$HR1[i:(i+5)])) < 0.1) {
    data4$HR1[i] = NA
  } 
}


for (j in seq(NROW(data4))){
  if (j+5 > length(seq(NROW(data4)))) {
    data4$HR2[j]=NA
  } else if (mean(data4$ID[j:(j+5)]) != data4$ID[j]) {
    data4$HR2[j] = NA
  } else if ((sd(data4$HR2[j:(j+5)])) < 0.1) {
    data4$HR2[j] = NA
  } 
}
```


## Now we need to run some analysis

Let's start with a multilevel model that accounts for 
- stability (how each signal is autocorrelated)
- interpersonal dependence (each signal is dependent from the previous state of the other signal)

The data needs to be further prepared, so we can analyze both participants in the same model.
We need to turn the data into a long format:
- a column indicating own hr and one own respiration
- a column indicating other hr and one other respiration
- a column indicating change in hr from previous round and one in respiration

We can then run an analysis where change is a function of one's previous state (stability, see slides), and the other's previous state (coupling). Make sure to:
- set up the most interesting contrasts: how do these parameters vary by condition? which condition should be baseline?
- set up the right random effects.
- N.B. the model will be slow. Make sure it works on a subset of the data first!

Bonus question: what if we include an additional layer? Is my heart rate just adjusting to yours, or also to how much you are adjusting to mine?
- to start answering this we can add a column indicating the previous change in hr in the other and one in respiration
- we can then build on the previous models by also adding the previous change in the other


```{r}
# Genearate a column for HR1_change and HR2_change
data4 <- data4 %>% 
  group_by(ID )%>% 
  mutate(HR1_change = lead(HR1)-HR1, HR2_change = lead(HR2)-HR2)

# Make long HR_other collumn
HR_other <- data4 %>%
  select(HR1, HR2, ID) %>% 
  gather("P", "HR_other", 2:1) %>% 
  ungroup() %>%
  select(HR_other)

HR_other <- as.data.frame(HR_other)

# Make long HR_change collumn
HR_change <- data4 %>%
  select(HR1_change, HR2_change, ID) %>% 
  gather("Person", "HR_change", 1:2) %>% 
  ungroup() %>%
  select(HR_change, Person) %>% 
  mutate(Person = as.factor(Person))

HR_change <- as.data.frame(HR_change)

# Making long HR_change_other collumn
HR_change_other <- data4 %>%
  select(HR1_change, HR2_change, ID) %>% 
  gather("Person", "HR_change_other",2:1) %>% 
  ungroup() %>%
  select(HR_change_other)

HR_change_other <- as.data.frame(HR_change_other)

# Making long HR_self collumn
model_data <- data4 %>%
  gather("Person", "HR_self", 3:4) %>% 
  select(time, study, group, trial, condition, ID, HR_self)


# Giving normal data label: type = 1 (needed for contrasting with surrogate pairs)
model_data <- as.data.frame(model_data)
model_data$type <- 1

# Binding all relevant collumns together
model_data <- cbind(model_data, HR_other, HR_change, HR_change_other)


# Making models testing effects of parameters for each condition seperately
model_1 <- lm(HR_change ~ 0 + condition + (HR_self+HR_other):condition, data=model_data)
summary(model_1)
plot_summs(model_1)


# Adding random effects
model_2 <- lmer(HR_change ~ 0 + condition + (HR_self+HR_other):condition + (1+HR_self| ID) + (1+HR_self| group), data=model_data)
summary(model_2)
summ(model_2, digits=3)



# Set the most interesting contrast e.g. by defining synchronous or conversation as the baseline
# Default baseline: Conversation
model_0 <- lmer(HR_change ~ 1 + (HR_self+HR_other)*condition + (1+HR_self| ID) + (1+HR_self| group), data=model_data)
summary(model_0)
summ(model_0, digits=3)
```


## Now we need to create control baselines.

First shuffled controls, then surrogate pairs.

### Creating controls: shuffled controls

Shuffled controls break the temporal dependencies of time-series by shuffling the value within one time-series. This ensures the "coordination" observed is not due to the actual values in the series and not their sequence.
Tip: sample() is your friend, but make sure to shuffle things within participant/condition and not throughout the whole dataset
 
```{r}

# Create a shuffled dataset
shuffled_data <- model_data %>% 
  group_by(ID, Person) %>% 
  mutate(HR_self = sample(HR_self), HR_other = sample(HR_other)) %>% 
  mutate(HR_change = lead(HR_self)-HR_self, HR_change_other = lead(HR_other)-HR_other, type = 0)

shuffled_data <- as.data.frame(shuffled_data)

# Concatenate it to the original dataset 
shuffled_model_data <- rbind(shuffled_data, model_data)
shuffled_model_data$type <- as.factor(shuffled_model_data$type)

# Create the same models as in the previous chunk, but adding an interaction by shuffled vs. real
model_3 <- lmerTest::lmer(HR_change ~ (HR_self+HR_other):condition:type + (1 | ID) +(1 | group), data=shuffled_model_data)
summary(model_3)
summ(model_3, digits=3)
```
 
 
### TRICKY! Creating controls: surrogate pair controls
 - Per each real pair, identify at least one surrogate pair (matching one of the participants, with somebody doing the same task, but in a different pair)

```{r}

# Identify unique pairs within a given study (to keep things manageable) and create list of all possible surrogate pairs 
Groups <- as.numeric(as.character(unique(model_data$group)))
SurrogateList <- expand.grid(a = Groups, b = Groups)
SurrogateList <- subset(SurrogateList, a != b)

surrogate_data <- data.frame()
data4 <- data4 %>% ungroup()

# Make data for all possible mixed, surrogate pairs
for (d in 1:nrow(SurrogateList)) {
  x <- subset(data4, group==SurrogateList$a[d])
  y <- subset(data4, group==SurrogateList$b[d])
  
  for (co in c("Synchronous","TurnTaking","SelfPaced","Conversation", "MovementGuided")) {
      
      z1 <- as.data.frame(subset(x,condition==co))
      z2 <- as.data.frame(subset(y,condition==co))
      z1 <- z1 %>% mutate(ID=paste(as.character(SurrogateList$a[d]),as.character(SurrogateList$b[d]),co, sep=""))
      
      if (NROW(z1) > NROW(z2)) {
        z1 <-  z1[c(1:nrow(z2)),]
      }
      if (NROW(z2) > NROW(z1)) {
        z2 <-  z2[c(1:nrow(z1)),]
        }
        
        w1 <- z1 %>% mutate(
          HR2 = z2$HR2,
          HR2_change = z2$HR2_change) 
        
        surrogate_data <- rbind(surrogate_data,w1)
  }}
      

# Making long formatc collumns of surrogate pair data
sur_other <- surrogate_data %>%
  select(HR1, HR2) %>% 
  gather("P", "HR_other", 2:1) %>% 
  select(HR_other)

sur_other <- as.data.frame(sur_other)

sur_change <- surrogate_data %>%
  select(HR1_change, HR2_change) %>% 
  gather("Person", "HR_change", 1:2) %>% 
  select(HR_change, Person) %>% 
  mutate(Person = as.factor(Person))

sur_change <- as.data.frame(sur_change)

sur_change_other <- surrogate_data %>%
  select(HR1_change, HR2_change) %>% 
  gather("Person", "HR_change_other",2:1) %>% 
  select(HR_change_other)

sur_change_other <- as.data.frame(sur_change_other)

sur_data <- surrogate_data %>%
  gather("Person", "HR_self", 3:4) %>% 
  select(time, study, group, trial, condition, ID, HR_self)


sur_data <- as.data.frame(sur_data)
# Giving surrogate pairs data label: type = 0
nsur_data$type <- 0

# Bind relevant collumns
sur_data <- cbind(sur_data, sur_other, sur_change, sur_change_other)

# Combine normal 
sur_model_data <- rbind(sur_data, model_data)
sur_model_data$type <- as.factor(sur_model_data$type)

# Running models contrasting surrogate-pairs results with normal data (adding'type' as interaction)
model_4 <- lmer(HR_change ~ 0 + condition + (HR_self+HR_other):condition:type + (1+HR_self| ID) + (1+HR_self| group), data=sur_model_data)
summary(model_4)
summ(model_4, digits=3)

# No random effects
model_5 <- lm(HR_change ~ 0 + condition + (HR_self+HR_other):condition:type, data=sur_model_data)
summary(model_5)
summ(model_5, digits=3)
```
 

### Effects of respiration coordination on heart rate coordination
 - describe how you would test those.
 - Optional: run the models and report them

 